---
title: "Beta Regression"
subtitle: "When the response is a proportion"
code-fold: show
warning: false
message: false
---

::: {.callout-note}
## About this page: an AI-first draft

As with the [Ordinal](../ordinal-regression/index.qmd) and [Negative
Binomial](../negative-binomial-regression/index.qmd) pages, this was drafted *first* by an
AI assistant (Claude Fable), to close a gap the [GLM Dashboard
Explainer](https://stats-board.jonminton.net/) had opened by running ahead
of these notes. The R here is deliberately built from base R — we *simulate* data and fit it
by hand — both so the page renders without extra packages and because that approach suits
the likelihood-and-simulation spirit of this course. Read the prose critically before
relying on it.
:::

## Where this fits

These notes keep returning to the two-part model (after King, Tomz & Wittenberg, 2000):

$$
Y_i \sim f(\theta_i, \alpha) \qquad\qquad \theta_i = g(X_i, \beta)
$$

Each model so far has been a particular choice of family $f$ and link $g$. Beta regression
is the choice for a response that is a **continuous proportion** — a number strictly between
0 and 1 — and it makes a point worth pausing on: it reuses the **exact same logit link as
logistic regression**, but pairs it with a *different distribution*. Link and family are
separate choices, and this is the cleanest demonstration of that in the whole course.

## The data: proportions, rates and shares

Plenty of responses live on the open interval $(0, 1)$ without being yes/no outcomes: the
proportion of exam questions a student gets right, a candidate's share of the vote in a
constituency, the fraction of an insurer's premiums paid out in claims, the proportion of a
budget spent. These are continuous, but bounded.

Two familiar models both misfit them:

- **Ordinary linear regression** treats the response as unbounded, so it will cheerfully
  predict a proportion of $1.3$ or $-0.2$, and it assumes a constant variance that a bounded
  quantity cannot have (a proportion near 0 or 1 simply has less room to vary).
- **Logistic regression** is for *binary* outcomes — counts of successes out of trials — not
  a continuous proportion that was never a count in the first place. Forcing a 0.73 to be
  "0.73 of a success" throws away the fact that it is a genuine measured fraction.

Beta regression keeps the response on $(0,1)$ and lets its variance shrink toward the
boundaries, where bounded data really do vary less.

## The beta distribution

The beta distribution lives on $(0, 1)$ and is usually written with two shape parameters,
but for regression it is far more natural to reparameterise it in terms of a **mean** $\mu$
and a **precision** $\phi$:

$$
\operatorname{E}(Y) = \mu, \qquad \operatorname{Var}(Y) = \frac{\mu(1-\mu)}{1 + \phi}.
$$

The mean $\mu$ is what the predictors will steer; $\phi$ plays the role that $\sigma^2$ does
for the Gaussian or $\theta$ does for the negative binomial — a dispersion-style parameter
that the data also have to estimate. Larger $\phi$ means *more* precise, i.e. less variance
around the mean. The distribution's great virtue is flexibility of shape while staying
inside the boundaries:

```{r}
#| fig-cap: "The beta distribution takes many shapes on (0,1) — here for three (mean, precision) pairs — but never strays outside the boundaries, unlike a Gaussian."
#| fig-width: 7
#| fig-height: 3.2
library(ggplot2)

shapes <- list("mu = 0.3, phi = 15" = c(0.3, 15),
               "mu = 0.5, phi = 8"  = c(0.5, 8),
               "mu = 0.7, phi = 25" = c(0.7, 25))
grid <- do.call(rbind, lapply(names(shapes), function(nm) {
  p <- shapes[[nm]]
  x <- seq(0.001, 0.999, length.out = 300)
  data.frame(x = x, density = dbeta(x, p[1] * p[2], (1 - p[1]) * p[2]), shape = nm)
}))

ggplot(grid, aes(x, density, colour = shape)) +
  geom_line(linewidth = 1) +
  labs(x = "y (a proportion)", y = "density", colour = NULL,
       title = "Beta densities on (0, 1)") +
  theme_minimal() +
  theme(legend.position = "top")
```

## The same logit link, a different distribution

To connect the mean to predictors we need a link that keeps $\mu$ inside $(0,1)$. That is
exactly the job the **logit** did for logistic regression, and we use it again unchanged:

$$
\mu_i = \operatorname{logit}^{-1}(X_i\beta) = \frac{1}{1 + e^{-X_i\beta}}.
$$

So the logit keeps reappearing across this course, each time bolted onto a different family:
**binomial** for binary outcomes (logistic regression, in [Intro to
GLMs](../intro-to-glms/index.qmd)); the **cumulative** form for ordered categories ([Ordinal
Regression](../ordinal-regression/index.qmd)); and now **beta** for continuous proportions.
The link handles the *constraint*; the family handles the *shape* of the randomness. They
are genuinely independent dials.

## Fitting it: simulate, then recover

Beta regression in R is usually done with the `betareg` package, which is not part of base R.
Rather than depend on it, we can lean on the [likelihood and simulation
theory](../likelihood-and-simulation-theory/index.qmd) developed earlier and do the whole
thing from scratch: invent data with a *known* answer, then let an optimiser climb the
log-likelihood back to it. If the machinery is sound, the estimates should land on the
truth.

First, simulate proportions from a beta model with coefficients we choose:

```{r}
set.seed(2024)
n  <- 300
x1 <- runif(n, -2, 2)
x2 <- rbinom(n, 1, 0.5)
X  <- cbind(1, x1, x2)

beta_true <- c(0.4, 0.8, -0.5)                 # intercept, x1, x2
phi_true  <- 15                                # precision
mu <- plogis(X %*% beta_true)                  # the logit link
y  <- rbeta(n, mu * phi_true, (1 - mu) * phi_true)
```

The beta log-likelihood for one observation, in the mean–precision parameterisation, is

$$
\ell_i = \ln\Gamma(\phi) - \ln\Gamma(\mu_i\phi) - \ln\Gamma\!\big((1-\mu_i)\phi\big)
       + (\mu_i\phi - 1)\ln y_i + \big((1-\mu_i)\phi - 1\big)\ln(1-y_i),
$$

which `dbeta(..., log = TRUE)` computes for us. We hand the negative of its sum to `optim`
— the "Robo-Chauffeur" of the likelihood page — optimising $\log\phi$ so the precision stays
positive:

```{r}
neg_loglik <- function(par) {
  beta <- par[1:3]
  phi  <- exp(par[4])                          # keep precision positive
  mu   <- plogis(X %*% beta)
  -sum(dbeta(y, mu * phi, (1 - mu) * phi, log = TRUE))
}

fit <- optim(c(0, 0, 0, 0), neg_loglik, method = "BFGS", hessian = TRUE)
se  <- sqrt(diag(solve(fit$hessian)))          # SEs from the Hessian, as on the likelihood page

data.frame(
  parameter = c("(Intercept)", "x1", "x2"),
  true      = beta_true,
  estimate  = round(fit$par[1:3], 3),
  std_error = round(se[1:3], 3)
)
```

The estimates sit right on top of the true values, and the precision is recovered too:

```{r}
c(true = phi_true, estimate = round(exp(fit$par[4]), 2))
```

Note the standard errors came straight from the curvature of the log-likelihood at its peak
— the inverse Hessian — exactly the connection between curvature and uncertainty made in the
[likelihood and simulation theory](../likelihood-and-simulation-theory/index.qmd) notes. The
same Hessian an optimiser uses to find the peak efficiently is the one that tells you how
sure to be once you are there.

## The one-liner in practice

In day-to-day work nobody hand-rolls the likelihood; the `betareg` package wraps all of the
above — and adds proper inference, a model for the precision, diagnostics and predictions —
into a single call that mirrors `glm`:

```{r}
#| eval: false
library(betareg)
m <- betareg(y ~ x1 + x2)        # same estimates as above, plus full inference
summary(m)
```

That `betareg` is a separate package, rather than another `family` inside base R's `glm()`,
is itself instructive: the GLM *framework* reaches comfortably past the handful of families
R ships with.

## Why only betas look at betas (again)

Because the link is the logit, the coefficients are once more on the **log-odds scale**, not
the proportion scale — so, as with logistic and ordinal regression, you cannot read a $\beta$
off and call it an effect on the proportion. Map it through the inverse logit for a concrete
profile, or read $\exp(\beta)$ as an odds ratio, but do not stare at the raw number.

And the precision parameter $\phi$ is worth a glance of its own. It is the beta family's
version of the negative binomial's $\theta$ and the Gaussian's $\sigma^2$: a second,
stochastic-side parameter that the data must estimate and that governs how tightly the
responses cluster around the fitted mean. Across these extended families, the same lesson
keeps surfacing — the $\alpha$ in $f(\theta, \alpha)$ is not always a nuisance to be ignored.

## Try it interactively

::: {.callout-tip}
## Try it interactively
The companion [GLM Dashboard
Explainer](https://stats-board.jonminton.net/) has an interactive beta
walkthrough — choosing a bounded-proportion response, seeing the logit link reused, and
watching the beta distribution's shape flex inside $(0,1)$: [Tutorial 6: Bounded
Proportions](https://stats-board.jonminton.net/tutorials/06-beta/systematic.html).
:::

## Further reading and connections

- **The link, reused.** [Intro to GLMs](../intro-to-glms/index.qmd) introduces the logit for
  logistic regression, and [Ordinal Regression](../ordinal-regression/index.qmd) reuses it
  in cumulative form — beta regression is the third appearance of the same link.
- **The machinery.** [Likelihood and Simulation
  Theory](../likelihood-and-simulation-theory/index.qmd) is where `optim`, the
  log-likelihood and the Hessian-as-uncertainty come from, and the [Complete Simulation
  Example](../complete-simulation-example/index.qmd) is the simulate-then-recover loop used
  above, in full.
- **The interactive companion.** The other extended families have tutorials on the [GLM
  Dashboard Explainer](https://stats-board.jonminton.net/).
