---
title: "Ordinal Regression"
subtitle: "When the response is ordered categories"
code-fold: show
warning: false
message: false
---

::: {.callout-note}
## About this page: an AI-first draft

This page was drafted *first* by an AI assistant (Claude Fable), and is offered as a
starting point for me to revise rather than as a finished, hand-checked piece.

It exists because of a small role-reversal. The written notes here are meant to be the
canonical home for the theory, with the [GLM Dashboard
Explainer](https://stats-board.jonminton.net/) acting as the interactive
companion. But while building out the dashboard, the companion ran *ahead* of the course:
the dashboard gained a complete ordinal-regression tutorial (and a few other model
families) while JonStats had no ordinal material at all. Rather than leave that gap, the
dashboard work was used to seed this page.

Treat the worked numbers below as machine-generated. The R code is real and runnable, but
the prose around it deserves a careful read before you rely on it.
:::

## Where ordinal regression fits

Throughout these notes the organising idea has been the two-part view of a model (after
King, Tomz & Wittenberg, 2000): a **stochastic component** saying what kind of random
thing the response is, and a **systematic component** saying how the predictors push it
around.

**Stochastic Component**

$$
Y_i \sim f(\theta_i, \alpha)
$$

**Systematic Component**

$$
\theta_i = g(X_i, \beta)
$$

Linear, logistic and Poisson regression are all just particular choices of $f$ (the
family) and $g$ (the link). Ordinal regression is another choice — but it is the first one
in this course where the *systematic* side has to stretch a little, because a single
intercept is no longer enough.

```{mermaid}
flowchart LR
  X
  beta
  eta
  thresholds
  Y
  X --> eta
  beta --> eta
  eta --> thresholds
  thresholds --> Y
```

The new ingredient, compared with the pipeline in [Intro to
GLMs](../intro-to-glms/index.qmd), is the `thresholds` box between the linear predictor and
the observed category. That box is the whole story of this page.

## The data: ordered categories

An **ordinal** response is one whose categories have a natural order, but no claim to equal
spacing. Survey items on a Likert scale (*strongly disagree* … *strongly agree*), pain
recorded as *none / mild / moderate / severe*, satisfaction as *low / medium / high* — all
ordered, none safely treated as numbers.

We'll use `MASS::housing`, the canonical ordinal dataset: a frequency table of
householders' satisfaction with their housing, cross-classified by the type of housing,
their sense of influence over management, and their degree of contact with other
residents.

```{r}
library(MASS)
data(housing)
head(housing)
```

The response `Sat` is an **ordered factor** with three levels, `Low < Medium < High`.

Two tempting shortcuts are both wrong, and wrong in instructive ways:

- **Treat the categories as numbers** (`Low = 1`, `Medium = 2`, `High = 3`) and run a
  linear regression. This smuggles in an assumption nobody actually believes: that the
  gap from *low* to *medium* equals the gap from *medium* to *high*. It will also happily
  predict a satisfaction of `2.7`, which means nothing.
- **Treat the categories as unordered** and run a multinomial model. This is safe but
  wasteful: it throws away the ordering, estimating a separate set of coefficients for
  every category and so spending far more parameters than the problem needs.

Ordinal regression threads between the two: it respects the ordering without pretending the
gaps are equal.

## A latent variable, cut into categories

The standard way to make sense of "ordered but not equally spaced" is to imagine a
**continuous latent variable** sitting behind the observed category — a quantity of
"satisfaction" that really is continuous, which we only ever see coarsened into *low*,
*medium* or *high* by where it falls relative to a set of **thresholds** (or cut-points).

Concretely, write the latent quantity for individual $i$ as a linear predictor plus noise,
$\eta_i = X_i\beta$, and place $J-1$ thresholds $\alpha_1 < \alpha_2 < \dots <
\alpha_{J-1}$ along its scale. We observe category $j$ when the latent value lands between
$\alpha_{j-1}$ and $\alpha_j$. With a logistic noise term this gives the **cumulative
logit** (or *proportional odds*) model:

$$
P(Y_i \le j) = \operatorname{logit}^{-1}(\alpha_j - X_i\beta), \qquad j = 1, \dots, J-1,
$$

with the conventions $P(Y_i \le 0) = 0$ and $P(Y_i \le J) = 1$. The probability of an
individual category is then the difference of two neighbouring cumulative curves,
$P(Y_i = j) = P(Y_i \le j) - P(Y_i \le j-1)$.

The crucial structural fact — the thing that makes ordinal regression *ordinal* — is that
there are **$J-1$ thresholds for $J$ categories**, and the same slope vector $\beta$ acts
at every threshold. The predictors move the whole latent distribution left or right; the
thresholds stay put and decide how that movement gets carved into categories.

::: {.callout-tip}
## The picture, hand-drawn

I've drawn this latent-variable-with-thresholds construction out by hand before, in a
slightly different setting: [Factor analysis with ordinal
variables](https://blog.jonminton.net/posts/handdrawn-stats/factor-analysis-ordinal-variables-analogue/),
in the hand-drawn statistics series on the blog. The diagram on page four of that post —
a continuous scale chopped into ordinal levels by a row of thresholds, with the rule "if
there are $T$ levels there are $T-1$ thresholds" — is *exactly* the construction above.

One honest caveat, because the two settings are not the same model. That post is about
**measurement**: recovering a latent *factor* from a battery of ordinal indicators (the
thresholds there are estimated per item, and there are no predictors). This page is about
**regression**: a single ordinal *outcome* whose latent scale is driven by predictors
$X\beta$, with one shared set of thresholds. Same atom — *which slice of an
unequally-spaced latent continuum do you fall into?* — put to two different uses. If you
want the static, measurement-flavoured intuition, read the drawing; if you want the
moving, predictor-driven version, read on (or open the [interactive
demo](#try-it-interactively)).
:::

## Fitting it in R

`MASS::polr` ("proportional odds logistic regression") fits the cumulative logit model.
The `housing` data come pre-tabulated, so we pass the cell counts as `weights`.

```{r}
m <- polr(Sat ~ Infl + Type + Cont, weights = Freq, data = housing, Hess = TRUE)
summary(m)
```

The output has two blocks, and the split matters. The **Coefficients** are the slopes
$\beta$ — one per predictor, *not* one per threshold. The **Intercepts** are the thresholds
$\alpha_1, \alpha_2$ — here labelled `Low|Medium` and `Medium|High`, the two cut-points
between three categories. There is no single "intercept": its job has been taken over by
the $J-1$ thresholds.

We can show the fitted thresholds on the latent scale, which is the rendered, real-data
version of the hand-drawn picture above:

```{r}
#| fig-cap: "The two fitted thresholds cut the latent (logistic) satisfaction scale into Low, Medium and High. The curve is the latent noise distribution for a household at the baseline of every predictor."
#| fig-width: 7
#| fig-height: 3.2
library(ggplot2)

thr <- as.numeric(m$zeta)                      # fitted thresholds (cut-points)
d <- data.frame(x = seq(-4, 4, length.out = 400))
d$density <- dlogis(d$x)

ggplot(d, aes(x, density)) +
  geom_area(fill = "grey85") +
  geom_vline(xintercept = thr, linetype = "dashed") +
  annotate("text", x = c(-1.5, 0.1, 1.9), y = 0.265,
           label = c("Low", "Medium", "High")) +
  labs(
    x = "latent satisfaction scale",
    y = "density",
    title = "J = 3 categories, J - 1 = 2 thresholds"
  ) +
  theme_minimal()
```

Move a household toward higher influence or away from terraced housing and you slide the
whole curve rightward under those fixed dashed lines, shifting probability mass from *Low*
toward *High*.

## Why only betas look at betas (again)

A recurring theme of these notes is that the raw $\beta$ coefficients are usually *not* the
quantity you want to read off and interpret directly. Ordinal regression is a clean case in
point: a coefficient here is neither a change in the response (the response is a category,
not a number) nor a probability. On the logit scale it is a **log odds ratio**, so the
interpretable quantity is its exponential:

```{r}
round(exp(coef(m)), 3)
```

Because the model is *proportional odds*, each of these has a single, threshold-independent
reading: $\exp(\beta_k)$ is the multiplicative effect on the odds of being in a **higher**
category rather than a lower one, for a one-unit change in $x_k$ — and that same factor
applies at *every* cut-point. So high influence over management multiplies the odds of
greater satisfaction by about $3.6$, while living in a terrace (versus a tower block)
multiplies them by about $0.34$.

That "one push across all thresholds" is precisely the assumption that lets ordinal
regression be so economical — and the thing you should be suspicious of (see below).

If you want something genuinely interpretable on the scale people care about, predict
probabilities for a concrete profile rather than staring at coefficients:

```{r}
newd <- data.frame(Infl = "Medium", Type = "Tower", Cont = "High")
round(predict(m, newd, type = "probs"), 3)
```

A medium-influence tower-block household with high resident contact is predicted *most*
likely to report high satisfaction — but with real probability mass in all three
categories, which is exactly the kind of honest, hedged statement a categorical outcome
deserves.

## The proportional-odds assumption

The economy of the model — one slope vector instead of one per threshold — rests entirely
on the proportional-odds assumption: that a predictor's effect is the *same* at every
cut-point. That is a real, testable restriction, and it is not always met.

A few practical notes:

- It can be checked formally (for example a Brant test, or by comparing against a model
  that lets the slopes vary across thresholds).
- The `ordinal` package's `clm()` is a more flexible modern alternative to `polr`, and
  supports *partial* proportional odds, where you relax the assumption for just the
  predictors that need it.
- A latent variable made of *normal* rather than *logistic* noise gives the **ordered
  probit** model — the same threshold construction, a different cumulative curve, and
  usually near-identical conclusions. This is the same logit-versus-probit choice that
  appears for binary outcomes in [Intro to GLMs](../intro-to-glms/index.qmd).

## Try it interactively

::: {.callout-tip}
## Try it interactively
The companion [GLM Dashboard
Explainer](https://stats-board.jonminton.net/) has a full ordinal-regression
walkthrough — choosing the response and predictors, picking the cumulative-logit link,
and an interactive version of the threshold picture above where you can drag the
latent distribution under the cut-points and watch the category probabilities change:
[Tutorial 7: Ordered
Categories](https://stats-board.jonminton.net/tutorials/07-ordinal/systematic.html).
:::

## Further reading and connections

- **The hand-drawn origin.** [Factor analysis with ordinal
  variables](https://blog.jonminton.net/posts/handdrawn-stats/factor-analysis-ordinal-variables-analogue/)
  gives the latent-variable-and-thresholds intuition in pen and paper, and its sequel
  [How factor analysis is used in
  testing](https://blog.jonminton.net/posts/handdrawn-stats/how-factor-analysis-is-used-in-testing/)
  carries the same idea into how standardised aptitude and knowledge tests are built —
  the bridge from ordinal models toward item-response theory.
- **The framework.** [Intro to GLMs](../intro-to-glms/index.qmd) sets up the
  stochastic/systematic split that this page extends, and the [why-only-betas
  argument](../intro-to-glms/index.qmd#why-only-betas-look-at-betas) that the odds-ratio
  interpretation above is an instance of.
- **The interactive companion.** Tutorials for the other model families — Poisson,
  negative binomial, gamma, beta and zero-inflated counts — live on the [GLM Dashboard
  Explainer](https://stats-board.jonminton.net/), several of which are
  not yet written up here.
