---
title: "Negative Binomial Regression"
subtitle: "When counts vary more than Poisson allows"
code-fold: show
warning: false
message: false
---

::: {.callout-note}
## About this page: an AI-first draft

Like the [Ordinal Regression](../ordinal-regression/index.qmd) page, this was drafted
*first* by an AI assistant (Claude Fable) and is offered as a starting point for me to
revise. It is part of closing a gap that opened when the [GLM Dashboard
Explainer](https://jonminton.github.io/glm-dashboard-explainer/) ran ahead of these notes:
the dashboard has a full negative-binomial tutorial, while JonStats had nothing on
overdispersion at all. The R code is real and runnable; the prose deserves a careful read.
:::

## Where this fits

The organising idea of these notes is the two-part model (after King, Tomz & Wittenberg,
2000): a **stochastic component** for what kind of random thing the response is, and a
**systematic component** for how the predictors move it.

$$
Y_i \sim f(\theta_i, \alpha) \qquad\qquad \theta_i = g(X_i, \beta)
$$

[Poisson regression](../intro-to-glms/index.qmd) is the standard first model for counts:
$f$ is the Poisson distribution and $g$ is the log link. But Poisson comes with a hidden
promise that real count data routinely breaks. Negative binomial regression keeps the same
systematic side and the same log link — it only changes $f$, and in doing so it puts the
usually-ignored parameter $\alpha$ to work.

```{mermaid}
flowchart LR
  X
  beta
  eta
  f
  alpha
  Y
  X --> eta
  beta --> eta
  eta --> f
  alpha --> f
  f --> Y
```

Everything on this page is about that $\alpha$ — the **dispersion** — flowing into $f$.

## The data: counts that won't sit still

We'll use `MASS::quine`: the number of days each of 146 Australian schoolchildren was
absent in a year, with predictors for ethnicity, sex, age group and learning ability.

```{r}
library(MASS)
data(quine)
c(mean = mean(quine$Days), variance = var(quine$Days))
```

The variance is roughly **sixteen times** the mean. That single fact is the whole problem,
because of a peculiarity of the Poisson distribution.

## Why Poisson isn't enough

The Poisson distribution has exactly one parameter, and it does double duty: the mean and
the variance are *the same number*,

$$
\operatorname{Var}(Y) = \operatorname{E}(Y) = \mu.
$$

That is a strong, testable claim — and counts in the wild rarely honour it. Real count data
are usually **overdispersed**: more variable than their mean, because of unmeasured
heterogeneity (some children are simply more absence-prone than their predictors reveal).

We can see the strain directly. If the Poisson model were adequate, its residual deviance
should be roughly equal to its residual degrees of freedom:

```{r}
m_pois <- glm(Days ~ Eth + Sex + Age + Lrn, family = poisson, data = quine)
c(resid_deviance = deviance(m_pois),
  df             = df.residual(m_pois),
  ratio          = deviance(m_pois) / df.residual(m_pois))
```

A ratio near twelve, where we wanted one, is the model telling us it cannot account for how
spread out the data are. The coefficients may not be badly wrong, but the *uncertainty* the
model reports will be — far too small, as we'll see.

## The negative binomial fix

The negative binomial keeps the mean model $\mu_i = \exp(X_i\beta)$ untouched and adds a
single **dispersion parameter** $\theta$ that lets the variance grow faster than the mean:

$$
\operatorname{Var}(Y) = \mu + \frac{\mu^2}{\theta}.
$$

This is a **variance function** — a statement about how spread depends on level — and it is
the first time in this course that the "uninteresting" stochastic parameter $\alpha$ has
become the point of the exercise. As $\theta \to \infty$ the extra term vanishes and the
negative binomial collapses back to the Poisson; small $\theta$ means heavy overdispersion.
Picturing the two variance functions side by side makes the difference concrete:

```{r}
#| fig-cap: "Poisson forces the variance to equal the mean (the straight line). The negative binomial lets it curve upward, with the fitted dispersion theta controlling how sharply."
#| fig-width: 7
#| fig-height: 3.4
library(ggplot2)
m_nb  <- glm.nb(Days ~ Eth + Sex + Age + Lrn, data = quine)
theta <- m_nb$theta
mu    <- seq(0, 60, length.out = 200)

vf <- rbind(
  data.frame(mu = mu, variance = mu,                  model = "Poisson:  Var = mu"),
  data.frame(mu = mu, variance = mu + mu^2 / theta,   model = "Negative binomial:  Var = mu + mu^2 / theta")
)

ggplot(vf, aes(mu, variance, colour = model)) +
  geom_line(linewidth = 1) +
  labs(x = "mean count (mu)", y = "variance", colour = NULL,
       title = "Two variance functions") +
  theme_minimal() +
  theme(legend.position = "top")
```

## Fitting it in R

`MASS::glm.nb` fits the negative binomial, estimating $\theta$ alongside the coefficients:

```{r}
summary(m_nb)
```

The coefficient block reads exactly like a Poisson fit — same log link, same kind of
log-rate coefficients. What is new is the line beneath it reporting the estimated
dispersion:

```{r}
m_nb$theta
```

A small $\theta$ like this confirms heavy overdispersion. And the model comparison is
decisive — the negative binomial is an enormously better fit for the same predictors:

```{r}
AIC(m_pois, m_nb)
```

## Why only betas look at betas (again)

A recurring lesson in these notes is that the raw coefficients are seldom the thing to read
off directly. With a log link they are log **rate ratios**, so the interpretable quantity is
their exponential:

```{r}
round(exp(coef(m_nb)), 3)
```

Each value multiplies the *expected count*. So switching the `Eth` indicator multiplies the
expected number of absences by about $0.57$ (roughly a 43% reduction), holding the other
predictors fixed.

But the deeper lesson here is about the numbers you *didn't* ask to see — the standard
errors. The Poisson and negative binomial models give nearly the same point estimates, yet
very different standard errors, because the Poisson model wrongly believes the data are far
tidier than they are. Fitting Poisson to overdispersed counts produces standard errors that
are too small, $t$-statistics that are too large, and a procession of significance stars
that should not be there. This is the count-data face of the warning in the [P-Values and
Statistical Significance](../../extra-courses/p-values-stat-sig/index.qmd) notes: a model
that misstates the variance will misstate your confidence, and overdispersion is one of the
commonest ways for that to happen quietly.

## Quasi-Poisson and other routes

The negative binomial is not the only response to overdispersion.

- **Quasi-Poisson** (`family = quasipoisson`) takes a blunter approach: it keeps the Poisson
  mean model but multiplies all the standard errors by an estimated dispersion factor. It
  fixes the inference without committing to a full distribution for the counts — useful, but
  it cannot be compared by AIC because it has no likelihood.
- **Zero-inflation** is a different problem again — an *excess of zeros* over what any single
  count distribution predicts, handled by a two-part mixture. The dashboard covers it as a
  separate tutorial; it stretches the framework further than this page does.

## Try it interactively

::: {.callout-tip}
## Try it interactively
The companion [GLM Dashboard
Explainer](https://jonminton.github.io/glm-dashboard-explainer/) has an interactive
negative-binomial walkthrough that lets you watch overdispersion break a Poisson fit and
then be absorbed by the dispersion parameter: [Tutorial 4: Overdispersed
Counts](https://jonminton.github.io/glm-dashboard-explainer/tutorials/04-negbin/systematic.html).
:::

## Further reading and connections

- **The framework.** [Intro to GLMs](../intro-to-glms/index.qmd) introduces the Poisson
  model and the log link that this page reuses unchanged.
- **The inference angle.** [P-Values and Statistical
  Significance](../../extra-courses/p-values-stat-sig/index.qmd) is the natural companion:
  overdispersion is a concrete reason a model's reported certainty can be untrustworthy.
- **The interactive companion.** The other extended families — gamma, beta, ordinal and
  zero-inflated counts — have tutorials on the [GLM Dashboard
  Explainer](https://jonminton.github.io/glm-dashboard-explainer/), some not yet written up
  here.
